### 2013-07-23 Tex/chapters/methods.tex
Gleichheitszeichen ergänzt
Fehlerkorrekturen

\subsection{Thermodynamic Theory}
\label{ssec:theory}
    To examine the disordered Ising model in Equilibrium, one can use
    methods of statistical physics. It is a canonical ensemble i.e. the
    temperature \(T\) is kept constant e.g. by a heat bath.
    In a canonical ensemble the probability \(p_i\) of a state
    \(i\) is distributed according to a Boltzman distribution
    \begin{equation}
        p_i \propto e^{-\beta H_i}
    \end{equation}
    \begin{equation}
        \beta = \frac{1}{k_B T}
    \end{equation}
    Further the free energy \(F\) of an canonical
    ensemble is minimized and all observables can be derived from \(F\)
    in a straight forward way, as stated in every textbook about
    statistical physics (e.g. \cite{nolting2005}).\\
    Because of \(F=U-TS\) where \(U=\avg{H}\) is the internal energy and
    \(S\) entropy, one can guess, that for low \(T\) the internal energy
    will be low, and for big \(T\) the entropy will be high, to minimize
    \(F\). The Hamiltonian of the Ising model creates an conflict between
    \(U\) and \(S\) by assigning low energies to states with high order.
    So \(U\) is low if all spins are aligned, where \(S\) as a measure of
    disorder is low. This preliminary considerations make a phase
    transition at some \(T\), where a high entropy lowers \(F\) more than
    a low internal energy, very plausible.\\\
    To determine
    \begin{equation}
        F=-k_{B}T \ln{Z}
    \end{equation}
    one has to know the partition function
    \begin{equation}
        Z = \sum_i p_i = \sum_i e^{-\beta H_i}
    \end{equation}
    where the sum goes over all possible states \(i\) of the system.
    Because every site can have two states, there are \(2^N\) different
    states of the system, which have to be summed up to gain \(Z\).
    Hence for small \(N\) the partition function is computable, but the system
    may show very different properties than in the thermodynamic limit.
    To minimize these finite size effects, it is desirable to examine
    systems with large \(N\). But \(2^N\) is a very rapidly increasing
    number, so there are next to infinity many different states, such
    that it is unfeasible to calculate the energy for each state, except
    for cases, where it is possible to solve it analytically.\\
    If not, one can get estimates of the observables for big \(N\) using
    Monte Carlo simulations.\\

\subsection{Monte Carlo Simulations}
\label{ssec:montecarlo}
    The idea behind Monte Carlo simulations is to take random samples of
    the observable, which should be measured, and estimate the mean of the observable from
    this samples. In example the Monte Carlo integration chooses random
    points within the bounding box of the function and estimates the integral
    as the fraction of points below the function and points above multiplied
    by the area of the bounding box \cite{Katzgraber2011}.\\
    Analogical the technique can be applied to statistical ensembles,
    but instead of sampling points on an area, one samples states of the
    system.\\
    In statistical physics the expected value of an observable \(O\)
    is calculated by
    \begin{equation}
        \avg{O} = \frac{1}{Z} \sum_i p_i O_i
    \end{equation}
    It is however possible, that there are few states that contribute
    massively more than others. Think of a canonical system at low \(T\),
    where \(p_i\) with low \(H_i\) contribute much more than high \(H_i\).
    But if one samples the \(2^N\) states evenly, it is probable to miss
    them, which possibly introduces big errors to the estimate of \(\avg{O}\).
    It is therefore desirable to sample only the states with high
    contributions to the sum.
    As noted before the system under scrutiny is canonical and therefore
    \(p_i\) is known. The states are distributed according to the Boltzman
    distribution. Hence \emph{Importance Sampling} can be utilized.\\
    Instead of sampling uniformly distributed random states, one samples
    states, which energies are distributed according to a Boltzman distribution.
    In fact this reduces the estimator to
    \begin{equation}
        O_M = \frac{1}{M} \sum_{i=0}^M O_i
    \end{equation}
    as shown in \cite{NewmanBarkema1999}. This is a very convenient form.\\
    But it is difficult to create a random state of a physical system e.g. the
    Ising system in equilibrium, according to a given distribution.
    Hence one uses \emph{Markov Chains} to generate new states \(\nu\)
    from old ones \(\mu\). It is important that the transition probabilities
    \(A(\mu \to \nu)\) obey \emph{Detailed Balance} and \emph{Ergodicity}.
    \emph{Detailed Balance} means that the probability to leave a state is
    the same as the probability to enter the state in equilibrium
    \(p_\mu A(\mu \to \nu) = p_\nu A(\nu \to \mu)\) with \(p_\mu\) the
    probability to be in state \(\mu\). This ensures that the system can
    equilibrate and that the states are distributed according to the
    desired distribution in equilibrium \cite{NewmanBarkema1999}.
    And \emph{Ergodicity} requires that every possible state is reachable
    from every other state in finite time. \cite{NewmanBarkema1999} \cite{Katzgraber2011}
    Otherwise the samples might not be representative for the whole system.\\
    In this thesis three algorithms, which fulfill all requirements,
    were used. They will be shortly described in the following subsections.

    \subsubsection{Metropolis}
        A Metropolis Monte Carlo\cite{Metropolis1953} simulation of an
        Ising model will choose a random spin, calculate the energy change
        \(\Delta H\) defined in eq. \eqref{eq:dH} that would result
        from a flip of that spin and execute the flip with the probability \(A\)
        eq. \eqref{eq:metropolis} \cite{NewmanBarkema1999} \cite{Katzgraber2011}.
        \begin{equation}
            \Delta H = H(\nu) - H(\mu)\\
            \label{eq:dH}
        \end{equation}
        \begin{equation}
            A(\mu \to \nu)
            \begin{cases}
                1                            & \Delta H \le 0 \\
                \exp{\brac{-\beta \Delta H}} & \Delta H > 0
            \end{cases}
            \label{eq:metropolis}
        \end{equation}
        So if a transition lowers the energy it is always done. This
        results in a high ratio between chosen spins and flipped spins.
        Therefore it minimizes the calculations needed for a change of
        the state.

>>>>

\subsection{Thermodynamic Theory}
\label{ssec:theory}
    To examine the disordered Ising model in Equilibrium, one can use
    methods of statistical physics. It is a canonical ensemble i.e. the
    temperature \(T\) is kept constant e.g. by a heat bath.
    In a canonical ensemble the probability \(p_i\) of a state
    \(i\) is distributed according to a Boltzman distribution
    \begin{equation}
        p_i \propto e^{-\beta H_i}
    \end{equation}
    \begin{equation}
        \beta = \frac{1}{k_B T}
    \end{equation}
    Further the free energy \(F\) of an canonical
    ensemble is minimized and all observables can be derived from \(F\)
    in a straight forward way, as stated in every textbook about
    statistical physics (e.g. \cite{nolting2005}).\\
    Because of \(F=U-TS\) where \(U=\avg{H}\) is the internal energy and
    \(S\) entropy, one can guess, that for low \(T\) the internal energy
    will be low, and for big \(T\) the entropy will be high, to minimize
    \(F\). The Hamiltonian of the Ising model creates an conflict between
    \(U\) and \(S\) by assigning low energies to states with high order.
    So \(U\) is low if all spins are aligned, where \(S\) as a measure of
    disorder is low. This preliminary considerations make a phase
    transition at some \(T\), where a high entropy lowers \(F\) more than
    a low internal energy, very plausible.\\
    To determine
    \begin{equation}
        F=-k_{B}T \ln{Z}
    \end{equation}
    one has to know the partition function
    \begin{equation}
        Z = \sum_i p_i = \sum_i e^{-\beta H_i}
    \end{equation}
    where the sum goes over all possible states \(i\) of the system.
    Because every site can have two states, there are \(2^N\) different
    states of the system, which have to be summed up to gain \(Z\).
    Hence for small \(N\) the partition function is computable, but the system
    may show very different properties than in the thermodynamic limit.
    To minimize these finite size effects, it is desirable to examine
    systems with large \(N\). But \(2^N\) is a very rapidly increasing
    number, so there are next to infinity many different states, such
    that it is unfeasible to calculate the energy for each state, except
    for cases, where it is possible to solve it analytically.\\
    If not, one can get estimates of the observables for big \(N\) using
    Monte Carlo simulations, which are introduced in the next chapter.\\
    %~ Once one has obtained values for e.g. the magnetization \(m\), one
    %~ can easily calculate the susceptibility
    %~ \begin{equation}
        %~ \chi = \frac{N}{T}\brac{\avg{m^2}-\avg{m}^2}
    %~ \end{equation}
    % Bisschen erklæren, dass man aus m auch chi machen kann und auf literatur verweisen

\subsection{Monte Carlo Simulations}
\label{ssec:montecarlo}
    The idea behind Monte Carlo simulations is to take random samples of
    the observable, which should be measured, and estimate the mean of the observable from
    this samples. In example the Monte Carlo integration chooses random
    points within the bounding box of the function and estimates the integral
    as the fraction of points below the function and points above multiplied
    by the area of the bounding box \cite{Katzgraber2011}.\\
    Analogical the technique can be applied to statistical ensembles,
    but instead of sampling points on an area, one samples states of the
    system.\\
    In statistical physics the expected value of an observable \(O\)
    is calculated by
    \begin{equation}
        \avg{O} = \frac{1}{Z} \sum_i p_i O_i
    \end{equation}
    It is however possible, that there are few states that contribute
    massively more than others. Think of a canonical system at low \(T\),
    where \(p_i\) with low \(H_i\) contribute much more than high \(H_i\).
    But if one samples the \(2^N\) states evenly, it is probable to miss
    them, which possibly introduces big errors to the estimate of \(\avg{O}\).
    It is therefore desirable to sample only the states with high
    contributions to the sum.
    As noted before the system under scrutiny is canonical and therefore
    \(p_i\) is known. The states are distributed according to the Boltzman
    distribution. Hence \emph{Importance Sampling} can be utilized.\\
    Instead of sampling uniformly distributed random states, one samples
    states, which energies are distributed according to a Boltzman distribution.
    In fact this reduces the estimator to
    \begin{equation}
        O_M = \frac{1}{M} \sum_{i=0}^M O_i
    \end{equation}
    as shown in \cite{NewmanBarkema1999}. This is a very convenient form.\\
    But it is difficult to create a random state of a physical system e.g. the
    Ising system in equilibrium according to a given distribution.
    Hence one uses \emph{Markov Chains} to generate new states \(\nu\)
    from old ones \(\mu\). It is important that the transition probabilities
    \(A(\mu \to \nu)\) obey \emph{Detailed Balance} and \emph{Ergodicity}.
    \emph{Detailed Balance} means that the probability to leave a state is
    the same as the probability to enter the state in equilibrium
    \(p_\mu A(\mu \to \nu) = p_\nu A(\nu \to \mu)\) with \(p_\mu\) the
    probability to be in state \(\mu\). This ensures that the system can
    equilibrate and that the states are distributed according to the
    desired distribution in equilibrium \cite{NewmanBarkema1999}.
    And \emph{Ergodicity} requires that every possible state is reachable
    from every other state in finite time. \cite{NewmanBarkema1999} \cite{Katzgraber2011}
    Otherwise the samples might not be representative for the whole system.\\
    In this thesis three algorithms, which fulfill all requirements,
    were used. They will be shortly described in the following subsections.

    \subsubsection{Metropolis}
        A Metropolis Monte Carlo\cite{Metropolis1953} simulation of an
        Ising model will choose a random spin, calculate the energy change
        \(\Delta H\) defined in eq. \eqref{eq:dH} that would result
        from a flip of that spin and execute the flip with the probability \(A\)
        eq. \eqref{eq:metropolis} \cite{NewmanBarkema1999} \cite{Katzgraber2011}.
        \begin{equation}
            \Delta H = H(\nu) - H(\mu)\\
            \label{eq:dH}
        \end{equation}
        \begin{equation}
            A(\mu \to \nu) =
            \begin{cases}
                1                            & \Delta H \le 0 \\
                \exp{\brac{-\beta \Delta H}} & \Delta H > 0
            \end{cases}
            \label{eq:metropolis}
        \end{equation}
        So if a transition lowers the energy it is always done. This
        results in a high ratio between chosen spins and flipped spins.
        Therefore it minimizes the calculations needed for a change of
        the state.

### 2013-07-23 Tex/chapters/results.tex
Fehlerkorrekturen

\subsection{Finite Size Effects}
\label{ssec:finitesize}
    The aim of this thesis is to find the critical temperature \(T_c\)
    of the disordered Ising model in dependence of the disorder parameter
    \(\sigma\). At \(T_c\) the mean absolute magnetization \(\avg{|m|}\) of
    the system will jump and the susceptibility
    \(\chi = \frac{N}{T}\brac{\avg{m^2}-\avg{m}^2}\)
    will diverge. In fig. \ref{fig:smeared_out}\subref{sfig:smeared_out:meanM_L}
    it is easy to see that the jump of \(\avg{|m|}\)
    occurs at lower temperatures \(T\) for higher
    disorder parameters \(\sigma\) (here with a RNG).
    \begin{figure}[htbp]
        \centering
        \subfigure[Dependence of the Phase Transition on $\sigma$][]
        {
            \label{sfig:smeared_out:meanM_L}
            \includegraphics[width=0.45\textwidth]{plots/Mean_M_L_128}
        }
        \subfigure[Example for Finite Size Effects][]
        {
            \label{sfig:smeared_out:meanM}
            \includegraphics[width=0.45\textwidth]{plots/meanM}
        }
        \caption[Phase Transition and Finite Size Effects]
        {
            \subref{sfig:smeared_out:meanM_L}: Effects of the disorder
            parameter $\sigma$ on the phase transition
            for an underlying RNG with $L=128$.\\
            \subref{sfig:smeared_out:meanM}: Effects of different system
            sizes at \(\sigma = 0\). Dotted lines are guides to the eye.
        }
        \label{fig:smeared_out}
    \end{figure}\\
    But in the shown plots there occurs no real jump, but a continuous
    decline. The jump is only present in infinite systems, hence no
    computer simulation will show the exact behavior in the
    thermodynamic limit. It will always show some \emph{finte size effects}.
    These finite size effects cause a "smearing out" of the phase
    transition. This is stronger for smaller system sizes, as is visible
    in fig. \ref{fig:smeared_out}\subref{sfig:smeared_out:meanM}\footnote{See the appendix \ref{appendix:finiteSizeEffects} for a similar figure for the specific heat.}.
    Clearly the \(L=16\) curve is much less steep than the \(L=128\) curve.\\
    Despite of this one can obtain \(T_c\) by finite size scaling
    methods \cite[S. ??]{NewmanBarkema1999} which also yield the critical
    exponents.

>>>>

\subsection{Finite Size Effects}
\label{ssec:finitesize}
    The aim of this thesis is to find the critical temperature \(T_c\)
    of the disordered Ising model in dependence of the disorder parameter
    \(\sigma\). At \(T_c\) the mean absolute magnetization \(\avg{|m|}\) of
    the system will show a steep decline to zero and the susceptibility
    \(\chi = \frac{N}{T}\brac{\avg{m^2}-\avg{m}^2}\)
    will diverge. In fig. \ref{fig:smeared_out}\subref{sfig:smeared_out:meanM_L}
    it is easy to see that the steep decline of \(\avg{|m|}\)
    occurs at lower temperatures \(T\) for higher
    disorder parameters \(\sigma\) (here with a RNG).
    \begin{figure}[htbp]
        \centering
        \subfigure[Dependence of the Phase Transition on $\sigma$][]
        {
            \label{sfig:smeared_out:meanM_L}
            \includegraphics[width=0.45\textwidth]{plots/Mean_M_L_128}
        }
        \subfigure[Example for Finite Size Effects][]
        {
            \label{sfig:smeared_out:meanM}
            \includegraphics[width=0.45\textwidth]{plots/meanM}
        }
        \caption[Phase Transition and Finite Size Effects]
        {
            \subref{sfig:smeared_out:meanM_L}: Effects of the disorder
            parameter $\sigma$ on the phase transition
            for an underlying RNG with $L=128$.\\
            \subref{sfig:smeared_out:meanM}: Effects of different system
            sizes at \(\sigma = 0\). Dotted lines are guides to the eye.
        }
        \label{fig:smeared_out}
    \end{figure}\\
    But in the shown plots there occurs no steep decline to zero, but a
    smooth decline. \(\avg{\abs{m}}(T_c) = 0\) is only present in infinite
    systems, hence no computer simulation will show the exact behavior in the
    thermodynamic limit. It will always show some \emph{finte size effects}.
    These finite size effects cause a "smearing out" of the phase
    transition. This is stronger for smaller system sizes, as is visible
    in fig. \ref{fig:smeared_out}\subref{sfig:smeared_out:meanM}\footnote{See the appendix \ref{appendix:finiteSizeEffects} for a similar figure for the specific heat.}.
    Clearly the \(L=16\) curve is much less steep than the \(L=128\) curve.\\
    Despite of this one can obtain \(T_c\) by finite size scaling
    methods \cite[S. ??]{NewmanBarkema1999} which also yield the critical
    exponents.
